Does differentiation beget differentiation?

### 2022-09-02 Manuscript_3_draft1.Rmd
Extensive revisions to Markdown manuscript, both in terms of code and content.

---
title: "Does differentiation beget differentiation?"
subtitle: Examining the link between implementation of and support for differentiation
thanks: "This is a very early draft. All comments are welcome at martin.moland@arena.uio.no. This version was finalized at `r format(Sys.time(), '%B %d, %Y')`."
author:
- name: Martin Moland
  affiliation: ARENA, Centre for European Studies
# bibliography: "`r paste0(Sys.getenv('HOME'),'/Dropbox/master.bib')`" # This would be more general
bibliography: inst/bib_paper_3.bib
csl: inst/european-journal-of-political-research.csl
abstract: "Questions about the public legitimacy of the EU’s increasingly flexible integration are attracting increasing scholarly interest. However, we still know little about whether exposure to past differentiation shapes public preferences for a more differentiated EU, and whether this varies according to how people think about EU membership. Combining recent survey data with data showing country-level exposure to differentiated integration, I show that those who have been exposed to more of it are likelier to want a future EU that allows countries to opt out of undesirable policy integration. This effect is particularly strong among Eurosceptics. My results suggest that exposure to the differentiated integration often used to integrate highly salient policy areas can solidify popular support for a less integrated EU, complicating the EU’s move towards an ‘ever closer Union’."
keywords: "differentiated integration, public opinion, Euroscepticism, European integration, contestation of the EU"
geometry: margin=1in
mainfont: cochineal
#sansfont: Linux Biolinum O
fontsize: 12pt
endnote: no
# pandocparas: TRUE
sansitup: FALSE
params:
  anonymous: ""
  doublespacing: ""
  removetitleabstract: ""
anonymous: "`r params$anonymous`"
doublespacing: "`r params$doublespacing`"
removetitleabstract: "`r params$removetitleabstract`"
header-includes:
  - \usepackage{longtable}
  - \LTcapwidth=.95\textwidth
  - \linespread{1.05}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \usepackage{setspace}\doublespacing
---

@@

# Introduction
Despite being among the world’s most advanced political cooperations, the expansion of the EU’s policy remit after the ratification of the Maastricht Treaty has made its integration ever-more differentiated [@leuffen_differentiated_2013; @schimmelfennig_ever_2020]. Internal EU differentiation typically takes two forms: First, some member states have opted out of cooperation to protect their own sovereignty. This is called “constitutional differentiation”. Second, through a process known as “instrumental differentiation”, countries with a diminished capacity for integration have been allowed to integrate at different schedules. This can happen either voluntarily or as a result of EU institutions imposing it as a precondition for membership [@schimmelfennig_instrumental_2014]. The differentiation of current European integration makes ‘variable geography’ [@stubb_categorization_1996] a better description of today’s EU than ‘ever closer Union’.

While increasing attention is being paid to why people want a differentiated EU [@leuffen_public_2022; @telle_elite_2022], we know little about how exposure to it has impacted support for a more differentiated future EU. While one of the few existing contributions examining the effect of differentiation on public opinion shows that it may strengthen the perceptions of the EU’s input legitimacy [@schraff_does_2020], two knowledge gaps remain: We still know too little about how past exposure to differentiation shapes support for it. Second, our understanding of how the relationship between exposure to differentiation and support for it is conditioned by attitudes towards the EU is still incomplete. Whether there is such a conditioning effect, either in the broader EU population, or where differentiation has been implemented in either a dominating or voluntary fashion is also unknown.

By combining survey data on citizens’ attitudes towards constitutional differentiated integration from 2020-2021 with data detailing differentiations from the EU’s treaties, I find that support for constitutional differentiation increases with exposure to differentiations from EU law, and that the extent to which this is the case strongly depends on your past attitudes towards the EU. However, I find that both where differentiation has been voluntary chosen, as in Scandinavia, or externally imposed in a discriminatory fashion, as in Central and Eastern Europe. While the reason for the generally greater support for differentiated integration among Eurosceptics in the countries that have been exposed to it may be that Eurosceptics are more likely than others to believe that differentiated integration is a path towards a less integrated Europe, more research is needed to understand why there is no real effect of pre-existing attitudes towards the EU in two regions with very different experiences of differentiation.

My findings are important to the discussion about the EU’s future spawned by Brexit [@european_commission_white_2017]: Because EU integration has expanded into increasingly salient areas, differentiation may be called upon to dampen public contestation of it [@schimmelfennig_ever_2020]. By providing greater numbers of citizens with exposure to differentiation, the EU’s institutions may solidify opposition to further integration among the Eurosceptics that have on several instances been able to put the brakes on further integration [@hooghe_postfunctionalist_2009].

@@

# Public support for European integration - what we know and what we don't
A large literature investigates individual preferences for European integration [@basile_sharing_2020; @bremer_junckers_2020; @de_vries_variable_2013; @eichenberg_post-maastricht_2007; @gabel_public_1998; @hobolt_public_2015; @hooghe_calculation_2005]. The literature identifies three mechanisms that help shape attitudes towards the EU. The first is utilitarian, with support for the EU contingent on whether individuals or social groups with which they identify stand to gain from integration [@foster_economic_2021; @gabel_public_1998]. The second is cue-based, and posits that citizens use heuristics such as elite discourses, media messages, and national benchmarks to form opinions about European integration  [@harteveld_europe_2013; @marquart_knowing_2019; @steenbergen_whos_2007; @vries_euroscepticism_2018]. The literature also shows that strong identification with the nation-state makes citizens more likely to oppose EU integration [@borzel_identity_2020; @hooghe_calculation_2005; @luedtke_european_2005].

An under-investigated issue is why some people support a more differentiated EU. This gap is particularly pronounced in light of the large that literature already investigates differentiated integration conceptually [@batora_towards_2020; @leuffen_differentiated_2013; @schimmelfennig_instrumental_2014], empirically [@ganzle_differentiated_2019; @malang_political_2020; @ott_between_2017; @schimmelfennig_eastern_2017; @winzen_government_2020] and normatively [@bellamy_republican_2019; @eriksen_contesting_2019; @lord_autonomy_2021; @nicolaidis_new_2004]. The few contributions investigating individual support for differentiated integration has focused on liberal economic values [@leuffen_public_2022], Euroscepticism [de_blok_blessing_2021], territorially exclusive self-identification [@moland_opting_2022] and party cues [@telle_elite_2022].

Thus, with one notable exception, national contextual factors are under-studied in this literature. However, these contexts are crucial for explaining attitudes towards both the EU and other international organizations [@dellmuth_elite_2021; @edwards_public_2009; harteveld_europe_2013]. One contextual source of support for differentiated integration may be the exposure that citizens of EU countries have had to it: A country’s number of differentiations from EU law may bring its level of integration in and out of alignment with what a citizen would prefer. If there is a misalignment between the actual and desired levels of integration, citizens may respond by adjusting their stated preference for future differentiated integration [@busemeyer_beyond_2021; @wlezien_public_1995]. While previous literature investigates the potential feedback loops between EU policy-making and public opinion [@bolstad_dynamics_2015; @de_bruycker_democratically_2020] this paper shifts the focus to both how support for EU differentiation is impacted by exposure to differentiated integration generally and whether this effect is conditioned by attitudes towards European integration.

@@

Responses to differentiated integration may further vary based on whether differentiation was imposed by the EU’s institutions or voluntarily chosen by member state government. An important reason for this is that citizens may perceive such differentiation as the EU dominating their countries, by shutting them out of economically beneficial integration, for instance in the field of freedom of movement [@cianciara_differentiated_2014; @schimmelfennig_instrumental_2014]. As @schimmelfennig_eu_2014 shows, the countries of Central and Eastern Europe were among those exposed to the greatest amount of discriminatory, rather than preferential, differentiated integration. The reason was that existing member states wanted to shield themselves from the economic destabilization believed to follow from the EU accession of countries like Poland, Estonia, and Hungary. As figure A6 shows, citizens of Central and Eastern Europe are among those in this sample exposed to the greatest amount of differentiated integration.

It is furthermore likely that Eurosceptics in these countries will have a more adverse reaction to differentiated integration than those who support EU membership: As differentiated integration in these countries has been predominantly discriminatory, they may be more likely to identify a non-uniformly integrated EU, in which countries can choose to opt out of integration if they so choose, with the potential for domination. Given the importance this group places on autonomy they are more likely to have adverse reactions to the discriminatory imposition of differentiated integration.

Elites in Central and Eastern Europe, especially in Poland, have also framed differentiated integration as a path towards a divided EU, with a powerful core and a less powerful periphery [@cianciara_differentiated_2014 178-179]. For example, Polish politicians opposed further integration of the Eurozone, a prominent example of differentiated integration, because they feared that the Central and Eastern member states, as non-members, would lose their influence over economic policy-making by not being inside the EU’s economic policy-making structures. Because elite cues are particularly important for shaping attitudes towards less salient issues, such as differentiated integration [@steenbergen_whos_2007; @telle_elite_2022], a reasonable assumption is that citizens would respond to these discourses by adjusting their preferences for integration downwards. Thus, two different causal mechanisms could produce the same outcome.

However, this history might be less relevant if it not being used by current actors. Instead, the increasingly Eurosceptic elite attitudes in Central and Eastern Europe [@borzel_identity_2020; @brack_towards_2020; @kriesi_politicization_2016] might form a more potent backdrop against which Central and Eastern European citizens might be expected to make up their minds. Exposure to such cues might, in contrast to what was stated before, make citizens more likely to express an increased demand for differentiated integration. This yields two competing hypotheses:

@@

```{r h2a, eval=!is_latex, results="asis"}
cat('> **H2a**: Exposure to differentiated integration in Central and Eastern will lead to lower levels of support for future differentiated integration than in the rest of the EU.')
```

*H2b: Exposure to differentiated integration in Central and Eastern Europe will lead to lower levels of support for it among Eurosceptics in these countries compared to supporters of EU membership.*

```{r h2b, eval=!is_latex, results="asis"}
cat('> **H2b**: Exposure to differentiated integration in Central and Eastern Europe will lead to lower levels of support for it among Eurosceptics in these countries compared to supporters of EU membership.')
```

*H3: Exposure to differentiated integration in Central and Eastern Europe will lead to higher levels of support for it than in the rest of the EU.*

```{r h3, eval=!is_latex, results="asis"}
cat('> **H3**: Exposure to differentiated integration in Central and Eastern Europe will lead to higher levels of support for it than in the rest of the EU.')
```

If exposure to discriminatorily imposed differentiation is assumed to lead to less support for differentiated integration, especially among Eurosceptics, the opposite is likely to be the case for countries with a great deal of voluntarily chosen differentiation. Scandinavia offers an interesting case for testing this assumption. Here the mode of implementation has been drastically different from that found in Central and Eastern Europe, as differentiation was both desired by national governments and justified through the autonomy-enhancing aspects of differentiated integration [@adler-nissen_opting_2014; @leruth_operationalizing_2015; @spendzharova_banking_2016]. Because of the previously explained cueing mechanism one might expect this pro-autonomy discourse to produce greater support for differentiated integration in Scandinavia than elsewhere in the EU.

@@

```{r h4a, eval=!is_latex, results="asis"}
cat('> **H4a**: The positive effect of exposure to differentiated integration will be greater for citizens of Sweden and Denmark than in the EU as a whole')
```

@@

```{r h4b, eval=!is_latex, results="asis"}
cat('> **H4b**: The positive effect of exposure to differentiated integration will be greater for Scandinavian Eurosceptics than supporters of EU membership.')
```

>>>>

---
title: "Does differentiation beget differentiation?"
subtitle: Examining the link between implementation of and support for differentiation
thanks: "This is a very early draft. All comments are welcome at martin.moland@arena.uio.no. This version was finalized at `r format(Sys.time(), '%B %d, %Y')`."
author:
- name: Martin Moland
  affiliation: ARENA, Centre for European Studies
# bibliography: "`r paste0(Sys.getenv('HOME'),'/Dropbox/master.bib')`" # This would be more general
bibliography: inst/bib_paper_3.bib
csl: inst/european-journal-of-political-research.csl
abstract: "Questions about the public legitimacy of the EU’s increasingly flexible integration are attracting increasing scholarly interest. However, we still know little about whether exposure to past differentiation shapes public preferences for a more differentiated EU, and whether this varies according to how people think about EU membership. Combining recent survey data with data showing country-level exposure to differentiated integration, I show that those who have been exposed to more differentiation are likelier to want a future EU that allows countries to opt out of undesirable policy integration. This effect is particularly strong among Eurosceptics. My results suggest that exposure to the differentiated integration often used to integrate highly salient policy areas can solidify popular support for a less integrated EU, complicating the EU’s move towards an ‘ever closer Union’."
keywords: "differentiated integration, public opinion, Euroscepticism, European integration, contestation of the EU"
geometry: margin=1in
mainfont: cochineal
#sansfont: Linux Biolinum O
fontsize: 12pt
endnote: no
# pandocparas: TRUE
sansitup: FALSE
params:
  anonymous: ""
  doublespacing: ""
  removetitleabstract: ""
anonymous: "`r params$anonymous`"
doublespacing: "`r params$doublespacing`"
removetitleabstract: "`r params$removetitleabstract`"
header-includes:
  - \usepackage{longtable}
  - \LTcapwidth=.95\textwidth
  - \linespread{1.05}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \usepackage{setspace}\doublespacing
---

@@

# Introduction
Despite being among the world’s most advanced political cooperations, the expansion of the EU’s policy remit after the ratification of the Maastricht Treaty has made its integration ever-more differentiated [@leuffen_differentiated_2013; @schimmelfennig_ever_2020]. Internal EU differentiation typically takes two forms: First, some member states have opted out of cooperation to protect their own sovereignty. This is called “constitutional differentiation”. Second, through a process known as “instrumental differentiation”, countries with a diminished capacity for integration have been allowed to integrate at different schedules. This can either happen voluntarily or as a result of EU institutions imposing it as a precondition for membership [@schimmelfennig_instrumental_2014]. The differentiation of current European integration makes ‘variable geography’ [@stubb_categorization_1996] a better description of today’s EU than ‘ever closer Union’.

While increasing attention is being paid to why people want a differentiated EU [@leuffen_public_2022; @telle_elite_2022], we know little about how exposure to it has impacted support for a differentiated future EU. While one of the few existing contributions examining the effect of differentiation on public opinion shows that it may strengthen the perceptions of the EU’s input legitimacy [@schraff_does_2020], two knowledge gaps remain: We still know too little about how past exposure to differentiation shapes support for it. Second, our understanding of how the relationship between exposure to differentiation and support for it is conditioned by attitudes towards the EU is still incomplete. Whether there is a conditioning effect, either in the broader EU population, or where differentiation has been implemented in either a dominating or voluntary fashion is also unknown.

By combining survey data on citizens’ attitudes towards constitutional differentiated integration from 2020-2021 with data detailing differentiations from the EU’s treaties, I find that support for constitutional differentiation increases with exposure to differentiations from EU law, and that the extent to which this is the case strongly depends on your past attitudes towards the EU. However, I find that both where differentiation has been voluntary chosen, as in Scandinavia, or externally imposed by the EU's institutions, as in Central and Eastern Europe, support for differentiated integration is higher than in the rest of the EU. While Eurosceptics may be more likely to support differentiated integration if they have been exposed to greater amounts of it because it would make them more likely than others to believe that differentiated integration is a path towards a less integrated Europe, more research is needed to understand why there is no real effect of past histories of differentiation in two regions with very different historical experiences with differentiated integration.

My findings are important to the discussion about the EU’s future spawned by Brexit [@european_commission_white_2017]: Because EU integration has expanded into increasingly salient areas, differentiation may be called upon to dampen public contestation of it [@schimmelfennig_ever_2020]. By providing greater numbers of citizens with exposure to differentiation, the EU’s institutions may solidify opposition to further integration among the same Eurosceptics that have previously been able to put the brakes on further integration [@hooghe_postfunctionalist_2009].

@@

# Public support for European integration - what we know and what we don't
A large literature investigates individual preferences for European integration [@basile_sharing_2020; @bremer_junckers_2020; @de_vries_variable_2013; @eichenberg_post-maastricht_2007; @gabel_public_1998; @hobolt_public_2015; @hooghe_calculation_2005]. The literature identifies three mechanisms that help shape attitudes towards the EU. The first is utilitarian, with support for the EU contingent on whether individuals or social groups with which they identify stand to gain from integration [@foster_economic_2021; @gabel_public_1998]. The second is cue-based, and posits that citizens use informational shortcuts like elite discourses, media messages, and national benchmarks to form an opinion about European integration  [@harteveld_europe_2013; @marquart_knowing_2019; @steenbergen_whos_2007; @vries_euroscepticism_2018]. The literature also shows that strong identification with the nation-state makes citizens more likely to oppose EU integration [@borzel_identity_2020; @hooghe_calculation_2005; @luedtke_european_2005].

An under-investigated issue is why some people support a more differentiated EU, in which the EU institutionally permits the divergence of the EU's legal order. Such differentiated integration is already comprehensively investigated conceptually [@batora_towards_2020; @leuffen_differentiated_2013; @schimmelfennig_instrumental_2014], empirically [@ganzle_differentiated_2019; @malang_political_2020; @ott_between_2017; @schimmelfennig_eastern_2017; @winzen_government_2020] and normatively [@bellamy_republican_2019; @eriksen_contesting_2019; @lord_autonomy_2021; @nicolaidis_new_2004]. The few contributions investigating individual support for differentiated integration has focused on liberal economic values [@leuffen_public_2022], Euroscepticism [de_blok_blessing_2021], territorially exclusive self-identification [@moland_opting_2022] and party cues [@telle_elite_2022].

Thus, with one notable exception, contextual factors have not been brought to the fore in studies of public support for differentiated integration. However, national contexts are crucial for explaining attitudes towards both the EU and other international organizations [@dellmuth_elite_2021; @edwards_public_2009; harteveld_europe_2013]. One contextual source of support for differentiated integration may be the exposure that citizens of EU countries have had to it: A country’s number of differentiations from EU law may bring its level of integration in and out of alignment with what a citizen would prefer. If there is a misalignment between the actual and desired levels of integration, citizens may respond by adjusting their stated preference for future differentiated integration [@busemeyer_beyond_2021; @wlezien_public_1995]. However, previous literature has primarily investigated the potential feedback loops between EU policy-making, such as directives and regulations, and public opinion [@bolstad_dynamics_2015; @de_bruycker_democratically_2020]. This paper shifts the focus to how alterations to the EU's structure, in the form of negotiated or imposed exemptions from the EU's treaties, shapes attitudes towards future preferences for the very same structure.

@@

Responses to differentiated integration may further vary based on whether differentiation was imposed by the EU’s institutions or voluntarily chosen by member state government. An important reason is that citizens may perceive such differentiation as the EU dominating their countries, by shutting them out of economically beneficial integration. As @schimmelfennig_eu_2014 shows, the countries of Central and Eastern Europe were among those exposed to the greatest amount of discriminatory, rather than preferential, differentiated integration. Among other things, the new Eastern member states were for several years excluded from the mobility rights afforded to Schengen citizens. The reason was that existing member states wanted to shield themselves from the economic destabilization believed to follow from the EU accession of countries like Poland, Estonia, and Hungary. As figure A6 shows, citizens of Central and Eastern Europe are among those in this sample exposed to the greatest amount of differentiated integration.

Eurosceptics in these countries may also have a more adverse reaction to differentiated integration than those who support EU membership: As differentiated integration in these countries has been predominantly discriminatory, they may be more likely to identify a non-uniformly integrated EU, in which countries can choose to opt out of integration if they so choose, with the potential for domination. Given the importance this group places on autonomy they are more likely to have adverse reactions to the discriminatory imposition of differentiated integration.

Elites in Central and Eastern Europe, especially in Poland, have also framed differentiated integration as a path towards a divided EU, with a powerful core and a less powerful periphery [@cianciara_differentiated_2014 178-179]. For example, Polish politicians opposed further integration of the Eurozone, a prominent example of differentiated integration, because they feared that the Central and Eastern member states outside the currency area would lose their influence over the EU's economic policy-making. Because elite cues are particularly important for shaping attitudes towards less salient issues, such as differentiated integration, [@steenbergen_whos_2007; @telle_elite_2022] citizens could be more likely to respond to these discourses by adjusting their preferences for integration downwards than if they were related to more salient policy outcomes. Thus, two different causal mechanisms could produce the same outcome.

However, this history might be less relevant if it not being discursively mobilized by current actors. Instead, the increasingly Eurosceptic elite attitudes in Central and Eastern Europe [@borzel_identity_2020; @brack_towards_2020; @kriesi_politicization_2016] might form a more potent backdrop against which Central and Eastern European citizens might be expected to make up their minds. Exposure to such cues might, in contrast to what was stated before, make citizens more likely to express an increased demand for differentiated integration. This yields two competing hypotheses:

@@

*H2b: Exposure to differentiated integration in Central and Eastern Europe will lead to lower levels of support for it among Eurosceptics in these countries compared to supporters of EU membership.*  
*H3: Exposure to differentiated integration in Central and Eastern Europe will lead to higher levels of support for it than in the rest of the EU.*

If exposure to discriminatory and externally imposed differentiation is assumed to lead to less support for differentiated integration, especially among Eurosceptics, the opposite is likely to be the case for countries where differentiation has primarily been voluntarily chosen. Scandinavia offers an interesting case for testing this assumption. Here the mode of implementation has been drastically different from that found in Central and Eastern Europe, as differentiation was both desired by national governments and justified through the autonomy-enhancing aspects of differentiated integration [@adler-nissen_opting_2014; @leruth_operationalizing_2015; @spendzharova_banking_2016]. Because of the previously explained cueing mechanism one might expect this pro-autonomy discourse to produce greater support for differentiated integration in Scandinavia than elsewhere in the EU.

@@

@@
